### 2013-10-26 src2/thesis.tex
update

In figure~\ref{fig:researchmatrix} I our work is placed in relation to other languages,
embedded languages and libraries for parallel programming. The systems are roughly divided 
into three groups, based on their level of abstraction. There are low-level languages, here 
I place languages that are imperative and C-like. In the middle layer I place languages 
that go a bit further, they have higher level abstractions and are more easily composable. 
In the highest level I place languages that completely abstract away from details of the 
hardware they run on. The division is not free from personal bias and the placement of languages
in boxes was not always easy.

The figure also divides the languages according to what hardware they support. There 
are CPU specific languages, GPU specific languages and languages that support or aspire 
to support CPU,GPU and accelerator (Larrabee) execution.

With Obsidian we try to target the sparsely occupied area of mid-level GPU programming. With 
CUDA, the programmer has full control of how to divide the computation amongst threads and 
blocks. Using Accelerate, Nikola and Thrust the programmer would use high-level patterns
without direct insight into how the application would be mapped onto threads and blocks of the 
GPU. Using Obsidian we want to bridge that gap by both being more composable than CUDA 
but give the programmer more control of what the computation will look like on the GPU. This 
is done by allowing the programmer to specify and compose thread-level and block-level code. 
Recently, the CUB library arrived that is built on a similar idea, that GPU programmers need to 
have block/thread level control to get maximum performance, while maintaining a higher level 
and more Thrust-like programmer interface~\citet{CUB}. Before CUB, and as far as I am aware, 
we were alone at targeting this level.

The ArBB and EmbArBB systems also provide a set of parallel primitives that the programmer
composes to build his application. I place the ArBB and EmbArBB in a box slightly lower than 
Accelerate since while ArBB/EmbArBB also have built in reduction primitives, just like Accelerate, 
they are not general primitives. Where Accelerate have a single higher order reduction 
operation, ArBB/EmbArBB have {\tt reduce\_add, reduce\_mul} and so on. Benefits of 
having ArBB embedded in Haskell compared to C++ are discussed in section~\ref{sec:EmbArBB}.
%I will go into 
%what benefits there are to having ArBB embedded in Haskell compared to C++ in

@@

\subsection{Parallelism}

\noindent\emph{DPH}\citet{DPH} \newline
\noindent\emph{D.A.Accelerate}\citet{ACCELERATEDAMP11} \newline
\noindent\emph{Nikola}\citet{NIKOLA}\newline
\noindent\emph{Repa}\citet{REPA}\newline
\noindent\emph{Meta-Repa}\citet{METAREPA}\newline
\noindent\emph{ArBB}\citet{ARBB2011}\newline
\noindent\emph{OpenCL}\citet{OpenCL}\newline
\noindent\emph{CUDA}\citet{CUDA}\newline
\noindent\emph{Microsoft Accelerator}\citet{ACCELERATOR}\newline
\noindent\emph{Copperhead}\citet{copperhead}\newline
\noindent\emph{Delite}\citet{DELITE}\newline
\noindent\emph{A monad for deterministic parallelism (R. Newton)} \citet{MonadPar} \newline
\noindent\emph{Nesl}\citet{NESL} \newline

@@

\noindent\emph{What contributions are made by Obsidian}

>>>>

\subsection{Languages for data-parallel programming}

In figure~\ref{fig:researchmatrix} our work is placed in relation to other languages,
embedded languages and libraries for parallel programming. The languages and libraries consider 
are data-parallel; languages for control parallelism or distributed message passing system are 
not considered. The systems are roughly divided 
into three groups, based on their level of abstraction. There are low-level languages, here 
I place languages that are imperative and C-like. In the middle layer I place languages 
that go a bit further, they have higher level abstractions and are more easily composable. 
In the highest level I place languages that completely abstract away from details of the 
hardware they run on. The division is not free from personal bias and the placement of languages
in boxes was not always easy.

The figure also divides the languages according to what hardware they support. There 
are CPU specific languages, GPU specific languages and languages that support or aspire 
to support CPU,GPU and accelerator (Larrabee, Xeon Phi) execution.

% ---------------------------------------------------------------------------
% Obsidian, Accelerate, Nikola, Thrust, CUB 
% ---------------------------------------------------------------------------
With Obsidian we try to target the sparsely occupied area of mid-level GPU programming. With 
CUDA~\citet{wwwcuda} and OpenCL~\citet{OpenCL}, the programmer has full control of how to divide 
the computation amongst threads and blocks. Using Accelerate~\citet{ACCELERATEDAMP11}, 
Nikola~\citet{NIKOLA} 
and Thrust~\citet{THRUST} the programmer would use 
high-level patterns without direct insight into how the application would be mapped onto 
threads and blocks of the GPU. Using Obsidian we want to bridge that gap by both being 
more composable than CUDA but give the programmer more control of what the computation will 
look like on the GPU. This is done by allowing the programmer to specify and compose 
thread-level and block-level code. Recently, the CUB library arrived that is built on a 
similar idea, that GPU programmers need to have block/thread level control to get maximum 
performance, while maintaining a higher level and more Thrust-like programmer 
interface~\citet{CUB}. Before CUB, and as far as I am aware, we were alone at targeting 
this level.

% ---------------------------------------------------------------------------
% ArBB EmbArBB
% ---------------------------------------------------------------------------
The ArBB and EmbArBB systems also provide a set of parallel primitives that the programmer
composes to build his application. I place the ArBB and EmbArBB in a box slightly lower than 
Accelerate since while ArBB/EmbArBB also have built in reduction primitives, just like Accelerate, 
they are not general primitives. Where Accelerate have a single higher order reduction 
operation, ArBB/EmbArBB have {\tt reduce\_add, reduce\_mul} and so on. Benefits of 
having ArBB embedded in Haskell compared to C++ are discussed in section~\ref{sec:EmbArBB}.
%I will go into 
%what benefits there are to having ArBB embedded in Haskell compared to C++ in

% ---------------------------------------------------------------------------
% Microsoft Accelerator
% ---------------------------------------------------------------------------
Microsoft Accelerator~\citet{ACCELERATOR}, is a language embedded in C\#. Accelerator has
a data-parallel arrays and a set of aggregate operations (operations on entire arrays). The 
operations on data-parallel arrays are JIT-compiled to GPU code. Accelerator can also 
compile to multithreaded CPU code and make use of SIMD units of modern 
CPUs~\citet{ACCELERATORCPU}; this makes the Accelerator system related to ArBB.

% ---------------------------------------------------------------------------
% Python systems
% ---------------------------------------------------------------------------
There are many embedded languages implemented in Python that makes use of GPUs 
to improve performance. One example is Copperhead~\citet{copperhead}. Copperhead 
supports flat and nested data-parallelism via primitives such as {\tt map} and 
{\tt reduce}.

% ---------------------------------------------------------------------------
% Repa, Meta-Repa 
% ---------------------------------------------------------------------------

Repa (Regular Shape-polymorphic Arrays)~\citet{REPA}, is a Haskell library for 
data-parallel programming. 
Repa uses the same array representation as Obsidian, in Repa called a delayed
array, which is a function from index to element. However, in Repa arrays have shape, 
they can be of any dimensionality and there shape-polymorphic functions on these. 
For example {\tt map} can be applied to an array of any shape.

Meta-Repa is a reimplementation of the Repa library using deep embedded language 
techniques. Meta-Repa programs build ASTs that are then compiled (using template 
Haskell) to Haskell code using low-level parts of the Repa library for its parallelism.
The Meta-Repa approach to implementing the Repa library 
gives inlining of operations for free, where Repa programs needs to be annotated 
with pragmas to ensure inlining. On the other hand, in Meta-Repa the programmer 
needs to explicitly state when something should not be inlined (using a {\tt force} primitive).
This is related to our work on EmbArBB that also uses an existing library (ArBB) 
for its parallelism. In reference~\citet{FPCDSL}, a case is made for this style 
of embedding prexisting DSLs.

% ---------------------------------------------------------------------------
% DPH, NESL 
% ---------------------------------------------------------------------------
Data-Parallel Haskell~\citet{DPH} and Nesl~\citet{NESL} are languages for 
nested data-parallelism. Both run on CPUs, but there is also work towards 
a GPU version of Nesl~\citet{NestedGPU}.

% ---------------------------------------------------------------------------
% Delite, Vertigo 
% ---------------------------------------------------------------------------

Delite~\citet{DELITE} is an infrastructure for implementation of domain 
specific embedded languages. There is an extensible internal representation (IR,AST); 
the programmer extends some base IR with domain specific constructs. The Delite system 
has been used to implement domain specific embedded languages for machine learning, rendering, 
physics and other domains. Delite, being a framework and the languages implemented using it 
very domains specific, was impossible to place in the matrix (figure~\ref{fig:researchmatrix}).

@@

%\subsection{Parallelism}

%\noindent\emph{DPH}\citet{DPH} \newline
%\noindent\emph{D.A.Accelerate}\citet{ACCELERATEDAMP11} \newline
%\noindent\emph{Nikola}\citet{NIKOLA}\newline
%\noindent\emph{Repa}\citet{REPA}\newline
%\noindent\emph{Meta-Repa}\citet{METAREPA}\newline
%\noindent\emph{ArBB}\citet{ARBB2011}\newline
%\noindent\emph{OpenCL}\citet{OpenCL}\newline
%\noindent\emph{CUDA}\citet{CUDA}\newline
%\noindent\emph{Microsoft Accelerator}\citet{ACCELERATOR}\newline
%\noindent\emph{Copperhead}\citet{copperhead}\newline
%\noindent\emph{Delite}\citet{DELITE}\newline
%\noindent\emph{A monad for deterministic parallelism (R. Newton)} \citet{MonadPar} \newline
%\noindent\emph{Nesl}\citet{NESL} \newline

%\subsection{Embedded domain specific languages}

@@

\noindent\emph{What contributions are made by Obsidian} 
Looking at the related work, we se that almost all high level GPU programming 
languages provide abstractions of parallel primitives, aggregate operations, 
map-folds-zipwiths, giving justification to  our investigation of GPU programming without 
these.
